Soundness Proofs for Iterative Deepening

### 2015-11-11 pldi-16/proof/convergence.tex
clean things up in convergence; clarify the xxx

%%% weird backwards-induction thingie i came up with at the airport that doesn't work
%\begin{lemma}
%	Given a fully-relevant interleaving
%	\[
%		\mathcal{I} = \{ ..., (\tai,\pai), \ppothers{\pai}, (t_{\alpha(i+1)},p_{\alpha(i+1)}), ... \}
%	\]
%	if $\pai$ is a data-race PP, and the all PPs in the interleaving prefix
%	\[
%		\mathcal{J} = \{ ..., (\tai \cup t_{\alpha(i+1)}',p_{\alpha(i+1)}'), \ppothers(\pai)' \}
%	\]
%	are reachable, then $\pai$ is reachable.
%	\label{lem:coalesce}
%\end{lemma}
%
%Note here that in $\mathcal{J}$, some instructions of $t_{\alpha(i+1)}$ may change because they were reordered with $\ppothers(\pai)$;
%i.e., $\ppinstr(\pai)$ may not necessarily be the only site of thread communication,
%and some other conflicts may cause $t_{\alpha(i+1)}$ to behave differently.
%Hence we have labelled it $t_{\alpha(i+1)}'$, and similar for thread $\alpha$'s next PP.
%Nevertheless, the first instruction of $t_{\alpha(i+1)}'$ must still be the same as $\ppinstr{\pai}$.

@@

Consider the alternate interleaving prefix:
\[
	\mathcal{J} = \{..., (\tai \cup \ppnext\pai', \paj), \ppothers\pai' \}
\]
where we have reordered $\ppnext\pai$ to before the execution of $\ppothers\pai$.
Here, $\paj$ is the first sync API PP in $\alpha$ after $\pai$ (as $p_{\alpha(i+1)}$ may be a not-yet-reachable data race PP),
and $\ppnext\pai'$ may include transitions of $\alpha$ beyond $\ppnext\pai$ itself.
%and for brevity, we will abbreviate the union of transitions as $\tak$.
Then, $\ppothers\pai'$ are the other threads' transitions from our target interleaving.
except they may be altered by the presence of some {\em other} data race in $\ppnext\pai'$ occurring after $\ppinstr\pai$.
Likewise, $\ppnext\pai'$ may be altered by some other data race in $\ppothers\pai'$.
The only certainty so far is that $\ppnext\pai'$ begins with $\ppinstr\pai$.

It would be straightforward to show that any other data races in $\ppnext\pai'$, which would be discovered in this interleaving,
could be reordered to after $\ppothers\pai'$,
eventually transforming $\ppnext\pai'$ to contain no conflicting memory accesses but $\ppinstr\pai$ itself.
Unfortunately, $\mathcal{J}$ is not necessarily reachable, as $\ppothers\pai'$ still includes any data-race PPs from $\ppothers\pai$,
which were ordered after $\pai$ in $\mathcal{I}$ and hence not covered by the inductive assumption.
Hence, we must perform the same ``coalescing'' for each thread in $\ppothers\pai'$ that we did for $\alpha$,
which we will abbreviate:
\[
	\mathcal{K} = \{..., (\tai \cup \ppnext\pai', \paj), \coalesce{\ppothers\pai'} \}
\]
For each other thread $\beta$, $\coalesce{\ppothers\pai'}$ will contain a single transition,
starting with the first instruction of $\beta$'s corresponding transition in $\ppothers\pai$,
and ending with the next sync API PP in $\beta$, which we'll call $\pbk$.
Again, the only certainly-executed instruction by any thread after $\tai$ is $\ppinstr\pai$.
It is even possible that neither $\ppinstr\pai$ nor any other instruction by thread $\alpha$ will conflict with any other thread,
until a different data race PP between two other threads is discovered,
as shown in Figure~\ref{fig:threethreads}.

It follows from the inductive hypothesis (which covers the prefix indicated by ``$...$''),
and from sync API PPs being reachable by definition,
that $\mathcal{I''}$ is reachable.

@@

Now, we show by induction that a data race on $\ppinstr\pai$ will be reachable.
We assume that a state space $\mathcal{S}$ containing $\mathcal{K}$,
plus $n$ unique data race PPs among $\ppnext\pai'$ and $\ppothers\pai$,
will be reachable.
(In the base case, the number of new data-race PPs is 0, and $\mathcal{K}$'s reachability is justified above.)
We must show that in this state space, if the $\ppinstr\pai$ data race is not reachable,
a new unique data-race PP nevertheless will be.

By the definition of fully-relevant interleaving, it must be possible for some other thread $\beta$ to execute an instruction that conflicts with $\ppinstr\pai$.
%
If this conflict is not observed in $\mathcal{S}$,
% XXX: Handwave? I'm not entirely convinced by this step. How does it justify skipping all the way to the first possible "reachable" data race among the chain of PPs needed to expose \beta's communication with \ppinstr\pai?
it means that some preceding thread communication (involving $\beta$ and any number of other threads, possibly including $\alpha$)
must exist to have suppressed it,
and that not all possible interleavings of that communication were tested in $\mathcal{S}$.
%
By the soundness of DPOR \cite{dpor}, if a program behaviour is possible by interleaving threads at the boundaries of
the given transitions, that interleaving will be tested.
By contrapositive\footnote{
Still constructive: again, going from $A \rightarrow B$ to $\neg B \rightarrow \neg A$.},
because it was not tested, a necessary site of interleaving must be in the middle of some transition, rather than at its boundaries.

@@

Hence either a data-race including $\ppinstr\pai$ will be identified,
or an infinite/nonterminating sequence of other unique data-race PPs will be identified,
but by the halting assumption, this would constitute an infinite loop bug, which is sufficient.
(Alternatively, for any program with a finitely-sized instruction listing, the number of unique instruction pairs is finite.)
This finishes our inner induction.
Hence $\pai$ is reachable, finishing our outer induction.
Hence all PPs in $\mathcal{I}$ are reachable.
\end{proof}

%%%%%%%%%%%%%%%%% this didn't work out
%We proceed by inducting on the PPs according to the order in which for each PP $\pai$, $\ppnext{\pai}$ appeared in the trace.
%Note that this is not necessarily the same as the order of the $\pai$s themselves in $\mathcal{I}$;
%i.e., it is the order of the racy instructions rather than the order of their immediately preceding preemptions\footnote{
%Why use this strange ordering, rather than simply inducting by the order of preemptions?
%Figure~\ref{fig:why-ppnext} shows why the first PP, ordered by earliest point of preemption, might not necessarily be the first reachable one.
%It is also true that the $\ppnext{\pai}$ order does not necessarily match the PPs' reachability dependence relation, but it is easier to cover the exception cases using this ordering rather than vice versa.}.
%
%\begin{figure}[t]
%	\caption{to do}
%	\label{fig:why-ppnext}
%\end{figure}
%%%% well, the way we do it, might not be the first reachable one either...
%
%\newcommand\pbj{\ensuremath{p_{\beta{}j}}}
%\newcommand\pgk{\ensuremath{p_{\gamma{}k}}}
%For both the base case and inductive step, we know (vacuously, for the former) that for each other PP $\pbj$ with $\ppnext{\pbj} \prec \ppnext{\pai} \in \mathcal{I}$, $\pbj$ is reachable.
%To show $\pai$ is reachable, the proof has two parts.
%First, we must handle any preceding data-race PPs $\pgk$ with $\ppnext{\pai} \prec \ppnext{\pgk}$.
%Second, we must show that when {\pai} is removed, a data-race including $\ppinstr{\pai}$ will be identified.
%%%%%%%%%%%%%%%%%%%

>>>>

@@

{\bf Coalescing not-yet-reachable data race PPs.}
Consider the alternate interleaving prefix:
\[
	\mathcal{J} = \{..., (\tai \cup \ppnext\pai', \paj), \ppothers\pai' \}
\]
where we have reordered $\ppnext\pai$ to before the execution of $\ppothers\pai$.
Here, $\paj$ is the first sync API PP in $\alpha$ after $\pai$ (as $p_{\alpha(i+1)}$ may be a not-yet-reachable data race PP),
and $\ppnext\pai'$ may include transitions of $\alpha$ beyond $\ppnext\pai$ itself.
%and for brevity, we will abbreviate the union of transitions as $\tak$.
Then, $\ppothers\pai'$ are the other threads' transitions from our target interleaving.
except they may be altered by the presence of some {\em other} data race in $\ppnext\pai'$ occurring after $\ppinstr\pai$.
Likewise, $\ppnext\pai'$ may be altered by some other data race in $\ppothers\pai'$.
The only certainty so far is that $\ppnext\pai'$ begins with $\ppinstr\pai$.

It would be straightforward to show that any other data races in $\ppnext\pai'$, which would be discovered in this interleaving,
could be reordered to after $\ppothers\pai'$,
eventually transforming $\ppnext\pai'$ to contain no conflicting memory accesses but $\ppinstr\pai$ itself.
Unfortunately, $\mathcal{J}$ is not necessarily reachable, as $\ppothers\pai'$ still includes any data-race PPs from $\ppothers\pai$,
which were ordered after $\pai$ in $\mathcal{I}$ and hence not covered by the inductive assumption.
Hence, we must perform the same ``coalescing'' for each thread in $\ppothers\pai'$ that we did for $\alpha$,
which we will abbreviate:
\[
	\mathcal{K} = \{..., (\tai \cup \ppnext\pai', \paj), \coalesce{\ppothers\pai'} \}
\]
For each other thread $\beta$, $\coalesce{\ppothers\pai'}$ will contain a single transition,
starting with the first instruction of $\beta$'s corresponding transition in $\ppothers\pai$,
and ending with the next sync API PP in $\beta$, which we'll call $\pbk$.
Again, the only certainly-executed instruction by $\alpha$ after $\tai$ is $\ppinstr\pai$.
It is even possible that neither $\ppinstr\pai$ nor any other instruction by thread $\alpha$ will conflict with any other thread,
until a different data race PP between two other threads is discovered,
as shown in Figure~\ref{fig:threethreads}.

It follows from the inductive hypothesis (which covers the prefix indicated by ``$...$''),
and from sync API PPs being reachable by definition,
that $\mathcal{K}$ is reachable.

@@

{\bf Finding the next reachable data-race PP.}
Now, we show by induction that a data race on $\ppinstr\pai$ will be reachable.
We assume that a state space $\mathcal{S}$ containing $\mathcal{K}$,
plus $n$ unique data race PPs among $\ppnext\pai'$ and $\ppothers\pai$,
will be reachable.
(In the base case, the number of new data-race PPs is 0, and $\mathcal{K}$'s reachability is justified above.)
We must show that in this state space, if the $\ppinstr\pai$ data race is not reachable,
a new unique data-race PP nevertheless will be.

By the definition of fully-relevant interleaving, it must be possible for some other thread $\beta$ to execute an instruction that conflicts with $\ppinstr\pai$.
%
If this conflict is not observed in $\mathcal{S}$,
it means that some preceding thread communication (involving $\beta$ and any number of other threads, possibly including $\alpha$)
must exist to have suppressed it,
and that not all possible interleavings of that communication were tested in $\mathcal{S}$.
%
By the soundness of DPOR \cite{dpor}, if a program behaviour is possible by interleaving threads at the boundaries of
the given transitions, that interleaving will be tested.
By contrapositive\footnote{
Still constructive: again, going from $A \rightarrow B$ to $\neg B \rightarrow \neg A$.},
% XXX: Handwave. This neglects the possiblity that TWO necessariy sites of interleaving are as such, and that they are dependent (both preemptions must occur before either data race can be identified).
% TODO: Prove that there must exist one single transition with next paragraph's property, without needing to preempt on any other such transition.
because it was not tested, a necessary site of interleaving must be in the middle of some transition, rather than at its boundaries.

@@

{\bf Reaching $\mathbf{p_{\alpha{}i}}$.}
Hence either a data-race with $\ppinstr\pai$ will be identified,
or an infinite/nonterminating sequence of other unique data-race PPs will be identified,
but by the halting assumption, this would constitute an infinite loop bug, which is sufficient.
(Alternatively, for any program with a finitely-sized instruction listing, the number of unique instruction pairs is finite.)
This finishes our inner induction.
Hence $\pai$ is reachable, finishing our outer induction.
Hence all PPs in $\mathcal{I}$ are reachable.
\end{proof}
